---
title: "Influence of Dietary Patterns on Health and Disease"
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

```{r}
```{r}
library(tidyverse)

# Load data
diet <- read_csv("./data/DietData.csv")
participants <- read_csv("./data/ParticipantData.csv")

# Join datasets on 'id'
full_data <- left_join(diet, participants, by = "id")


```
```{r}


# Filter and round BMI
full_data <- full_data %>%
  filter(!is.na(bmi_adults), bmi_adults >= 12, bmi_adults <= 40) %>%
  mutate(bmi_round = floor(bmi_adults))

# Summarize nutrient energy per person and compute percentages
bmi_macros <- full_data %>%
  group_by(id, bmi_round) %>%
  summarise(
    total_energy = sum(energy, na.rm = TRUE),
    carb_energy = sum(carb * 4, na.rm = TRUE),
    fat_energy = sum(totalfat * 9, na.rm = TRUE),
    protein_energy = sum(totalpro * 4, na.rm = TRUE),
    fiber_energy = sum(fiber * 2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    carb_pct = carb_energy / total_energy * 100,
    fat_pct = fat_energy / total_energy * 100,
    protein_pct = protein_energy / total_energy * 100,
    fiber_pct = fiber_energy / total_energy * 100
  ) %>%
  select(bmi_round, carb_pct, fat_pct, protein_pct, fiber_pct) %>%
  pivot_longer(
    cols = c(carb_pct, fat_pct, protein_pct, fiber_pct),
    names_to = "macronutrient",
    values_to = "percent_kcal"
  ) %>%
  mutate(macronutrient = factor(macronutrient,
                                levels = c("fiber_pct", "protein_pct", "fat_pct", "carb_pct"),
                                labels = c("Fiber", "Protein", "Fat", "Carbohydrate"))) %>%
  group_by(bmi_round, macronutrient) %>%
  summarise(avg_percent = mean(percent_kcal, na.rm = TRUE), .groups = "drop") %>%
  arrange(bmi_round, macronutrient) %>%
  group_by(bmi_round) %>%
  mutate(
    cum_percent = cumsum(avg_percent),
    lower_bound = lag(cum_percent, default = 0)
  ) %>%
  ungroup()

# Plot
p <- ggplot(bmi_macros, aes(x = bmi_round, y = avg_percent, fill = macronutrient)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(12, 40, by = 2)) +
  labs(
    title = "Macronutrient Energy Contribution Across BMI Range",
    subtitle = "Stacked bars show average percent of energy from fiber, protein, fat, and carbohydrates (BMI 12–40)",
    x = "BMI",
    y = "Percent of Total Energy (%)",
    fill = "Macronutrient",
    caption = "Source: Global Dietary Database (https://globaldietarydatabase.org/management/microdata-surveys/653)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Add horizontal lines at each macronutrient boundary
for (nutrient in levels(bmi_macros$macronutrient)) {
  line_data <- bmi_macros %>%
    filter(macronutrient == nutrient) %>%
    select(bmi_round, lower_bound)
  
  p <- p + geom_line(
    data = line_data,
    aes(x = bmi_round, y = lower_bound),
    inherit.aes = FALSE,
    color = "black",
    size = 0.6
  )
}

# Add top boundary line at 100%
p <- p + geom_line(
  data = tibble(bmi_round = unique(bmi_macros$bmi_round), y = 100),
  aes(x = bmi_round, y = y),
  inherit.aes = FALSE,
  color = "black",
  size = 0.6
)

# Display plot
p


```

How do dietary patterns influence health and disease? Using data from NNMB survey collected from 2009 - 2012 for nutritional status of rural population,  we seek to examine the relationship between diet and prelevance of lifestyle diseases among adults. Diet patterns impact health and can pay a major role in life-style diseases like blood-pressue, diabetes and obesity. 

 